Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-13 paper/paper.tex
dwh put in distance prior note
updated example plot
some of bovys comments

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at all luminosities with useful parallax information.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all of the data available, fairly and responsibly.

@@

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.pdf}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper-left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper-right and lower-left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower-left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the expectation value of the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower-right panel shows the expectation value, and $1\sigma$ uncertainties, of the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value. The blue points are better than the black for doing stuff.}
\label{fig:toy}
\end{figure}

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong, (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas,
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof).
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical-Bayes, or maximum-marginalized-likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1 through 4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}

>>>>

However, the use of physical models for distance estimation is not necessary.
It is possible to build a stellar photometric model from the data themselves,
because there are stars at a wide range of luminosities with useful parallax information, and this
range of stellar luminosities with useful parallax information will grow with future \gaia\ data releases.
One challenge is that different stars are observed at different levels of parallax precision,
so it requires relatively sophisticated technology to build this data-driven model
using all of the data available, fairly and responsibly.

@@

\section{Why does this work?}
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{toy.pdf}
\caption{ {\bf Why does this work}: An example of generating posterior beliefs of true values from noisy data by using the noise-deconvolved distribution of data as a prior. The upper-left panel visualizes the simplified toy distribution. The red lines are the $1\sigma$ contours of the distribution, and the red points are 1024 samples from the distribution. These same $1\sigma$ contours are drawn in the bottom left panel. In the upper-right and lower-left panel, the grey points are these same 1024 samples of the distribution but with some measurement uncertainty added to them. The error bars show the $1\sigma$ measurement uncertainties. The black points are 5 randomly chosen data to highlight the method. In the lower-left panel, the blue lines represent the noise deconvolved, best estimate for the underlying distribution. The blue points represent the expectation value of the posterior belief of the true values of the noisy black points using the noise deconcolved distribution as the prior. The red points represent the actual true values. The lower-right panel shows the expectation value, and $1\sigma$ uncertainties, of the posterior belief of the true value for all the points. They are colored by the measurement uncertainty, showing that data with larger measured variance are more influenced by the prior and lie towards the center of the noise deconvolved distribution, or prior. In contrast, more certain measurements tend to lie closer to their measured value.}
\label{fig:toy}
\end{figure}

@@

To infer the true parallax for a star using the noise-deconvolved \cmd\ as
a prior, we make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the discussion section below.
\begin{description}
\item[stationarity] We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.
  In detail, this assumption is wrong. For example, more distant stars will have lower signal to noise ratio measurements
  and also will be different than local
  disk stars, in age and metallicity. The stationarity assumption is fairly weak:
  The requirement is that, locally in the \cmd, there is
  support among the high signal-to-noise stars for the types
  of stars also seen at low signal-to-noise.
  If this requirement is met, the results will not be
  strongly biased. Although this is a weak assumption, it is deep and fundamental to this
  work.
\item[selection] We have no model for the selection function of \tgas (but see \cite{bovy17}),
  nor selection volumes for any kinds of stars.
  For this reason, we are building a model of \emph{the contents of \tgas}, not the
  properties of the stars in the Milky Way (nor any volume-limited component or region
  thereof). For this reason, our \cmd\ prior might not be appropriate to use with other surveys. 
\item[big data] We assume that we have large numbers of stars, large enough that
  empirical-Bayes, or maximum-marginalized-likelihood, is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case, if a portion of the \cmd\ has only one (statistically distinct)
  exemplar---for example, a white dwarf---inference for these cases will be biased.
\item[noise model] We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the parallax uncertainty estimates,
  (the \texttt{parallax{\_}error} entry in the \gaia\ catalog), are correct.
  Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative, and leads to under-deconvolution. In other words, the features of the inferred \cmd\ will typically be broader and more conservatively estimated than than if we over-estimated the noise.
\item[dust] We treat the \psone-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical Bayes model, we iteratively update each star's
  dust correction at its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for neither
  color nor absolute magnitude.
\item[mixture of Gaussians] We assume that the \cmd\ can be represented by a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128 Gaussians.
\item[no physics] We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact).
\end{description}

@@

DWH: write something about how our prior is a prior on distance

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.pdf}
\caption{Five example posterior PDFs over parallax using our method.
The upper left panel shows the position of each example data point in black, with its observed uncertainties, plotted over our prior represented by a mixture of Gaussians in green.
We chose example points that evenly spanned the color space, but within each color bin, we chose a point at random.
The remaining panels show the posterior PDF in blue, the prior PDF in green, and likelihood of the data in black.
Most stars have posterior PDFs (blue) that are only slightly more precise than the data (black), which is well represented in this random sampling of points.
Examples 1 through 4 are not very effected by the prior (green), where good data is good data.
Example 4 is interesting because it lies in the same color range as the red clump, where there is a lot of support from the prior, but the likelihood does not have much support there so the parallax posterior PDF, transformed into an absolute magnitude, remains on the main sequence.
Example 5 shows a significantly more precise posterior PDF compared with the likelihood, and looks very similar to the prior PDF.
It also exemplifies that our prior tends to have the largest effect on stars in the red giant branch.}
\label{fig:posterior}
\end{figure}
